---
title: "Class05: Data Viz with ggplot"
author: "Emily Chase (PID: A14656894)"
format: gfm
---

Today we are playing with (gg)-plotting and graphics. 

There are *loads* of ways to make cool figures in R.

- There is "base" R graphics (`plot()`, `hist()`, `boxplot()`, etc)

- There are also add-on packages like **ggplot**

```{r}
# cars --> prints the entire thing when rendered
head(cars, 5)
```

## Let's plot this with "base" R:

```{r}

plot(cars)

```

Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp)
```


```{r}
hist(mtcars$mpg)
```


##  Now: ggplot

The main function in the ggplot2 package is `ggplot()`. First step if never used before is to install via `install.package(<package_you_want>)`. You can put this in quarto but every time you render it'll install the software, so you should do it in the R console. I've already installed this so now I just want to load it in.

```{r}
library(ggplot2)
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() 
```

Every ggplot needs
- **data** layer
- **aesthetic** layer
- **geom** layer

## Adding more layers

Add a line?

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() +
  geom_line()
```

...What about a fit line?

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() +
  geom_smooth()
```



```{r}
cars
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() +
  labs(title="Speed vs Stopping Distances of Cars",
       x="Speed (MPH)",
       y="Stopping Distance (ft)",
       subtitle="Linear relationship",
       caption="Dataset: 'cars'")   +
  geom_smooth(method="lm", se=FALSE) +
  theme_classic()
```


## Now we will plot some expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
> Q1 How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset

```{r}
nrow(genes)
```

> Q2 How many columns and what are their names?

```{r}
colnames(genes)
ncol(genes)
```
> Q3 How many up-regulated genes are there?

```{r}
sum(genes$State == "up") # returns logicals in the df, then sum it

# another way to get the number of occurences of things in a vector
table(genes$State)
```

> Q4 What percentage of genes are upregulated?

```{r}
tot_up = sum(genes$State == "up")
pct_up = tot_up/nrow(genes) * 100
print(pct_up)
```

> Plot the up/down regulated genes

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, color=State) +
  geom_point()
```

```{r}
p <- ggplot(genes) +
  aes(Condition1, Condition2, color=State) +
  geom_point() + 
  scale_color_manual(values=c("blue", "grey", "red")) +
  theme_bw()
p
```


```{r}
p+labs(title="Pretty rna seq data") 
```

## What if we added labels in aesthetic mappings?

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, color=State, label=Gene) +
  geom_point() + 
  scale_color_manual(values=c("blue", "grey", "red")) +
  theme_bw() +
  geom_text()
```

Yoinks


### Let's install a package that will help with labels
`install.packages("ggrepel")`

```{r}
library(ggrepel)
```



```{r}
ggplot(genes) +
  aes(Condition1, Condition2, color=State, label=Gene) +
  geom_point() + 
  scale_color_manual(values=c("blue", "grey", "red")) +
  theme_bw() +
  geom_text_repel(max.overlaps = 50)
```


## Making panels:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```


```{r}
tail(gapminder)
```


```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5) +
  facet_wrap(~continent) +
  theme_bw()
```


```{r}
ggplot(gapminder) +
  aes(lifeExp, fill=continent, size=pop) +
  geom_density(alpha=0.5, color="pink") +
  theme_classic() +
  labs(title="YUH")
```


